---
title: "Seminar Paper in Natural Resource Use and Conservation Economics: Preferences on Biodiversity Conservation"
author: "Tilman von Samson-Himmelstjerna (216232989 - MLU Halle-Wittenberg)"
date: "2023-02-06"
output:
  pdf_document: default
  word_document: default
reference-section-title: null
bibliography: lib.bib
linestrech: 1.5
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=FALSE)

```

```{r, include=FALSE, echo=FALSE, eval=TRUE}
source("startup.R")
source("dataadjust.R")
source("models.R")
source("mwtp.R")
```

## 1. Introduction

Biodiversity conservation is one of the most urging responsibilities for
modern societies to ensure a living within the planetary boundaries
(@steffen2015planetary). A sharp decline in the planet's biodiversity
has irreversible consequences not only for the stability of ecosystems,
but also for the stability of global economies. Recently, the corona
pandemic has highlighted the vulnerability of these systems and the
interconnections between human activities and the natural environment
(@lawler2021covid). A key challenge for biodiversity conservation
policies is the right incorporation of it's direct and indirect
values(@diaz2015ipbes). Decision-making must therefore be based on sound
economic valuation that captures the diversity of values and utilities
that people derive from the ecosystem and its biodiversity in
particular, which in most cases are not included in market prices.

To tackle this issue, a growing body of literature investigates on
stated preferences for biodiversity over the whole range of low-,
middle- and high-income countries (@bhat2021willingness,
@bienabe2006public, @le2016public). A cornerstone of this research is
the estimation of the respondents Willingness to Pay (WTP) for
biodiversity conservation. @bhat2021willingness, for instance, find a
high willingness to pay of around 3.32\$ per year in the form of
increasing water rates for biodiversity conservation in Dachigam
National Park, India. Findings like these can have far reaching
practical implications for local authorities and the design of
preference-based policy measures.

This choice experiment grounds on data that was collected between March
and April 2013 in Germany and aims at providing insights on preferences
regarding biodiversity conservation in a high-income country. Besides
the estimation of the respondents preferences on forest biodiversity and
4 other environmental attributes and the corresponding WTP, this study
aims at investigating on socio-demographic and behavioral factors that
might influence preferences with regard to biodiversity and thereby
could add further perspectives to the existing body of literature. On
the one hand it asks whether income might explain some of the effects of
preferences for biodiversity. The generic hypothesis would be, that
higher levels of income could leverage an individuals preferences for
natural goods as basic needs become less important. On the other hand
the study focuses on the effects of experiencing natural services by
spending days outside in the open landscape. For the latter the
hypothesis would follow the line that experiencing environmental
attributes might yield direct utility and an increase in the level of
biodiversity attributes might increase utility.

I find significant and positive estimates for preferences on the share
of forest, the share of unused forest and forest biodiversity, while the
estimate for the share of foreign trees is significant and negative. A
central finding is the negative impact of higher levels of income on
preferences on forest biodiversity.

Section 2. introduces the Econometric approach and provides some general
information on the data. While section 3 provides the results of the
model estimations and some interpretation, section 4 discusses these and
concludes.

## 2. Method

In order to estimate the respondents preferences over certain
environmental attributes I set up a discrete choice experiment that
builds up on discrete choice research and data on local land use changes
in Germany and especially on the work of @sagebiel2017 (referred to as
"the original paper"). Therefore, 1233 randomly German adults were
interviewed by a market research company between March and April 2013.
The respondents were given a set of choices that were framed as an
survey in order to learn more about people's views with regard to the
landscape, that was expressed in terms of six attributes. These included
*share of forest in the landscape*, *size of individual fields and
forest areas*, *share of foreign trees*, *share of unused forest*,
*biodiversity within the forest* and *price of improvement* on certain
attributes. Within each choice situation the respondents could choose
between three alternatives that reflected either improvement,
deterioration or persistence of the condition of the landscape with
regard to the specific attributes. The first level for the *share of
forest in the landscape* was a 10% decrease, the second the maintenance
of the status quo and the third an 10% increase in the share, which
correspondingly would lead to a proportional shift in the share of
agricultural land. Attribute levels for *biodiversity within the forest*
ranged between "as today", "slightly raised" and "distinctly raised".
The other attribute-levels followed similar patterns. The *price*
attribute was framed as a yearly payment to a local fund in order to
that ranged between 10€ and 160€.

Two important notes on the structure of the choices should be noted.
First, the data set needs to be adjusted for the individual status quo
levels, which was done in the original paper by including additional
data for the "share of forest" for instance. However, individual data on
the level of biodiversity was not available, which is important for my
analysis and must be considered by evaluating the results. Second, the
structure of the survey reflects mostly a discrete choice situation in
which the variables are ordinal, but not continuous and thus discrete.
In the original survey the levels of some of the attributes were not
ordered linearly. Therefore, I re-ordered them to estimate them in line
with the utility function specification, which is derived in the
following.

### Random utility model

The key underlying assumption of the choice experiment is that by
deciding for one of the alternatives, the respondents reveal their
preferences of the attributes which then can be expressed in a indirect
utility function and ultimately be estimated within a conditional logit
model. Utility theory would imply that utility maximizing agents would
choose the alternative that yields the highest utility
(@louviere2000stated). The exact description of people's preferences
within a linear utility function is an unrealistic assumption when it
comes to estimating real world preferences, as almost certainly
randomness is involved in the decision context. Therefore, I include the
random parameter Epsilon to the econometric model. The deterministic
part from the choice observations is expressed in terms of an indirect
utility function. The decision towards the higher valued alternative
reflects the solution of the individuals maximization problem and can be
captured within the indirect utility function. However, uncertainty
demands that conclusions must be drawn in terms of probabilities, due to
the unobserved part Epsilon. Therefore the choice results can be
understood such that the probability of choosing one alternative over
the other is the probability that the utility of one alternative is
bigger than the other. The response probabilities then can be derived by
using a conditional logit model, which follows a cumulative distribution
function and is based on the random utility model which is derived in
the next part. In order to estimate the model I used Maximum Likelihood
Estimation.

### Model specification

With respect to the land use choice experiment I specified three
different models. Each model is based on a different utility function
that is expressed in terms of the six attributes that are modeled
differently each time.

In order to compare the different models I set up a baseline model with
a linear utility function:

$$U=\beta_0*ASC +\beta_1*ShFor+\beta_2*FiSizHalf+\beta_3*FiSizDouble+\beta_4*FoTre+$$
$$\beta_5*NoUse+\beta_6*FoBio +\beta_7*Pri+\epsilon$$

The attribute for the size of individual fields is coded across all
models as two dummy variables for each alternative. The second model is
very similar to the baseline model in the original paper by
@sagebiel2017. It contains squared terms for *ShFor*, as it is assumed
that the marginal utility of forest shares is diminishing. However, it
excludes the interaction term between the price attribute and the
disposable income:

$$U=\beta_0*ASC+\beta_1*ShFor+\beta_2*ShFor^2+\beta_2*FiSizHalf+\beta_3*FiSizDouble+$$
$$\beta_4*FoTre+\beta_5*NoUse+\beta_6*FoBio+\beta_7*Pri+\epsilon$$

The third model includes two interaction terms that aimed at the
initially postulated research question. The first is between the
biodiversity attribute and a mean-centered variable for disposable
income, the second is build between the biodiversity attribute and a
mean-centered variable for days in the open landscape:

$$U=\beta_0*ASC+\beta_1*ShFor+\beta_2*ShFor^2+\beta_3*FiSizHalf+\beta_4*FiSizDouble+\beta_5*FoTre+$$
$$\beta_6*NoUse+\beta_7*FoBio+\beta_8*Pri+\beta_9*{FoBio*MCInc}+\beta_10*{FoBio*MCNatDay}+\epsilon$$

The first interaction aims at capturing possible heterogeneity in
preferences that is explained by income. The hypothesis follows the
line, that higher income levels might increase the degree of
environmental preferences, as other needs, like basic needs, might
become relatively less important. However, as the data only captures
income on the county level, this can only be seen as a broad
approximation. As described above, the hypothesis for the latter
interaction states that utility for people that are spending more days
in the open landscape would increase by increasing levels of forest
biodiversity.

### Marginal Willingness to Pay (MWTP)

In the next step I calculated the marginal willingness to pay (MWTP)
with respect to the biodiversity attribute. The MWTP relates the
marginal utility of the attribute with the marginal utility of the cost
attribute and thus should reflect the monetary amount an individual
would pay for a marginal change in the level of the respective
attribute. The general formula for the MWTP is given by:

$$MWTP=-\frac{\frac{\partial V}{\partial atr}}{\frac{\partial V}{\partial a_c}}$$

This approach can be very practical in order to asses values for certain
goods or attributes that do not have a directly accessible market value
like the degree of biodiversity in the forest. In order to account for
the average willingness to pay one needs to account for a mean-centered
variant of the demographic variable that is used in the interaction
term. The average MWTP for the degree of biodiversity is calculated as:

$$MWTP_{biodiv}=-\dfrac{\beta_6+\beta_8*MCInc+\beta_9*MCNatDay}{\beta_7}$$

## 3. Results

The results are estimated with the Maximum Likelihood estimation by
using the "R" software package "Apollo" and are shown in Table 1. Unless
otherwise highlighted, the results for model three are discussed in the
following. The coefficients for the alternative specific constants (ASC)
are negative and do significantly vary from zero at the 0.1% level. As
the ASC's are coded as dummy variables they capture effects to not
choose the status quo option, that are not displayed by the attributes.
In both cases the estimates indicate decreases in utility.

The positive and significant coefficient at the 0.1% level for the
linear term of *ShFor* is large in magnitude across all model
specifications. It indicates that people gain utility in marginal
increases of forest shares. It can be assumed that with higher levels of
forest shares people gain less from an additional unit of forest. By
including a squared term for forest share, the model should account for
the possibility of diminishing marginal utility with respect to this
attribute. As the coefficient "b_ShFor2" is negative and the null
hypothesis can be refused at the 0.1% level, this assumption seems to be
justified and the hypothesis of an inversely U-shaped can be verified.

A less and a more fragmented landscape might not to be preferred by the
participants of the choice experiment. Both estimated dummy parameters
for the attribute *FiSiz* are negative. However, the coefficients are
not significant across all model specifications. This might imply that
the respondents prefer the status quo level over doubling and halving
the field and forest area sizes, but reasonable interpretation is not
feasible.

The parameter for *share of unused forest* is statistically significant
at the 0.1% level, large in magnitude and positive across all model
specifications. This implies that individuals gain utility with marginal
increases in unused forest shares. This insight might reflect the
intensive public discussion about the German "Waldsterben" and calls for
a more extensive forest management in the 1980s (@metzger2015erst). The
opposite effect can be observed with respect to the parameter *FoTre*
that captures increases in the amount of foreign trees, which is
negative and statistically significant at the 1% level.

The coefficient for *FoBio* is positive and significant, which implies
that on average the utility of biodiversity increases with higher levels
of biodiversity. Furthermore, the coefficient for the interaction term
between the biodiversity attribute and the mean centered average income
is significant at the 1% level and negative. In fact, this refuses the
initial hypothesis that higher levels of income lead to higher
preferences for biodiversity. The interaction between *days in the open
landscape* and biodiversity in the forest is neither large in magnitude,
nor statistically significant, which makes reasonable interpretation of
the result impossible despite the insight that within this discrete
choice setting an impact of experiencing "open landscape" directly has
no impact on preferences about biodiversity levels within the forest.

*Price* has a negative coefficient that is significant at the 0.1%
level. This indicates a decrease in utility with increasing costs for
improvement measures on the respective environment.

With a look at the two interaction terms it is remarkable that their
impact on the model outcomes is relatively small. By comparing the
results to the second model, which is termed as the "No Interaction"
model Table 2, one can see that further controls on socio-demographic
and behavioral variables do not change the model results significantly.
However, the coefficient for the income related interaction is
considerable in terms of magnitude, this effect would have been
unobserved by neglecting this variable. The addition of the second
interaction term does not add any explanatory power to the model and
could be dropped. However, in terms of the initially stated hypothesis
the result remains of interest. 
The main difference between the baseline
model and the subsequent models is the inclusion of the quadratic term for *ShFor*,
which has an considerable impact on the magnitude of the models
estimates. By comparing the baseline with the "No Interaction" model one
can see large differences for the the *ShFor* estimates, as its value drops from 0.43 to 0.07. The
decrease in the *ShFor* estimate and a reduction in the standard
errors shows that this utility function specification can capture the
dynamics and explain the data better than the linear specification. All 
estimates remain relatively similar when comparing the "No Interaction"
model and the "Full Model". A quick look at the Log-Likelihood values
indicates a better goodness-of-fit of the third model specification,
which can simply be explained by the higher number of predictor
variables. A more insightful indicator is the Likelihood ratio test, for
which the null hypothesis states that the nested model fits the data
equally well as the full model. The resulting Chi-squared test-statistic
is around 25.31 with a corresponding p-value at 3.193e-06, whereby the
null hypothesis that both models fit the data equally well can be
rejected at the 0.1% level. Therefore, including the interaction terms
adds more explanatory power to the model.

```{r, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
# lrtest(model3, model2)
# summary(lrtest)
```

```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
 
models_tex=list()

models_tex[["model1"]] <- quicktexregapollo(model1)
models_tex[["model2"]] <- quicktexregapollo(model2)
models_tex[["model3"]] <- quicktexregapollo(model3)
# screenreg(models_tex, digits=4)
```

```{r, echo=FALSE, results="asis", include=TRUE}
texreg(models_tex ,custom.model.names = c("Base Model", "No Interaction" , "Full Model"), custom.title="Table 1", caption = "Model Specifications")
```

### MWTP

A good way to find a practical interpretation of the results of a
discrete choice experiment is the calculation of the MWTP, which was
exemplary derived for *biodiversity in the forest* in the previous
section. Table 2 shows the MWTP for each attribute. Most of the MWTP
estimations for the attributes are significant at the 0.1% level. On
average the willingness to pay of the respondents for an marginal
increase of *FoBio* is 33.92€. The interpretation of the value is more
relevant in terms of the general direction of the value, then in precise
monetary units or units of biodiversity. As the scale of biodiversity
levels ranged from "as today" over "slightly increased" to "distinctly
increased" it is impossible to pin down a precise unit of biodiversity
that the MWTP would cover. It rather reflects the willingness to pay for
a subjectively perceived slight or distinct increase. Furthermore, the
data set was not adjusted for the individual specific status quo as data
for this attribute was unavailable. Therefore, only a general ordering
of preferences with regard to this attribute is covered by the data and
reflected in the MWTP value.

The MWTP for *ShFor* is positive but declining with further shares of
forest. This can be seen due to the positive value around 8.86€ that the
respondents would pay on average for an additional 1% share of forest
and the negative value around -0.06€ for the quadratic term that
indicates diminishing marginal utilities. Both terms are significant at
the 0.1% level.

```{r, echo=FALSE, results="asis", include=TRUE}

modeltex <- quicktexregapollo(model3,wtpest = allwtp)
# screenreg(modeltex, digits=4)
texreg(modeltex, digits=4, custom.model.names = "Full Model",  custom.title="Table 2",  caption = "Marginal Willingness to Pay")

```

## 4. Discussion and conclusion

In the context of the United Nations decade on biodiversity in the years
of 2011-2020 this study offers some insights on possible behavioral and
socio-demographic determinants of individual preferences for
biodiversity. Especially in the face of the ongoing increase of
anthropocentric driven biodiversity loss it is essential to determine
its drivers and possible angles for policy measures
{@jaureguiberry2022direct}. Whether or not contributions towards
ecosystem services lead to gains in individuals utility, has profound
policy implications.

In general, the Maximum Likelihood estimates of the conditional logit
model that is applied to the discrete choice experimental data in this
study are largely significant at the 0.01% level. With regard to forest
biodiversity the estimates indicate strong utility gains towards higher
degrees of biodiversity. In order to get more insights of these effects
I controlled for average regional income as a proxy for personal income
and for a variable that captures days spend in the open landscape. The
results show that higher levels of income lead to negative changes in
marginal utility of biodiversity. Therefore, the initial hypothesis that
income has positive impacts on biodiversity preferences needs to be
rejected. This finding is at odds with research done by
@bienabe2006public in a middle-income country, where the choice
experiment revealed that higher income groups have a higher WTP for
Payments for Ecosystem Services that enhance biodiversity conservation.
However, overall MWTP for increases in biodiversity levels is positive
and indicate on average a WTP around 34€ annually for slight increases
in forest biodiversity. This finding has important policy implications,
as taxes aiming on forest biodiversity protection could thereby be
justified. For example, a unit tax on the volume of timber harvested
would be conceivable (@michanek2018landscape). However, further research
would be necessary to verify the findings here with personal income data
and more precise choice-levels in order to find solid estimates for the
magnitude of possible policy interventions.

Experiencing days in the open landscape had no significant effect on the
respondents preferences on forest biodiversity. The empirical finding,
that direct experience and thus "usage" of biodiversity has no
significant impact on biodiversity preferences could theoretically
indicate the role of non-use "existence values". Nevertheless, a
reasonable determination of this relationship cannot be provided here
and could be an interesting trace for further research.

A possible limitation to these findings is the weak representatives of
the sample. @sagebiel2017 report that the respondents are on average
younger, less female, more educated and live in smaller households than
the German average. Another limitation is the lack of data on the
individual status quo of levels of forest biodiversity and personal
income data, which prevents the determination of exact estimates.
Furthermore, the estimation of the conditional logit model prevents from
controlling for preference heterogeneity, whereby, a mixed logit
approach could solve this issue. The large sample size of the survey and
the high significance of the results, nonetheless, can provide strong
inferences about the research question as initially formulated.

## Bibliography
